---
title: "MolecularEvoPhy"
author: "Yuejiao"
date: "2024-10-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(AnnotationForge)
library(clusterProfiler)
library(tidyverse)
library(ape)
library(phangorn)
library(factoextra)
library(ggplot2)
library(TreeDist)
library(cowplot)
library(ggpubr)
library(ComplexHeatmap)

```


## pair-wised species-tree vs gene tree
```{r}
## import the species tree
species_tree <- read.tree("/Users/hsf378/PhD/Analysis/turtle-jeans/Data/concord.cf.tree_concat.nwk")

## import the gene tree
locus.trees <- read.tree("/Users/hsf378/PhD/Analysis/turtle-jeans/Data/genetrees.nwk")

## list of genes included in this study
og_list <- read.table("/Users/hsf378/PhD/Analysis/turtle-jeans/Data/list.txt")
# names <- read.table("Data/list.txt")
# plot(species_tree)
# table(Ntip(locus.trees))

### calculate the pariwise normalized RF distance 
normal_rf <- c()
for (i in seq(length(locus.trees))) {
  rf <- TreeDist::RobinsonFoulds(species_tree,locus.trees[i],normalize = T)
  normal_rf <- c(normal_rf,rf)
}

names(normal_rf) <- og_list$V1

## Histogram of RF distance
hist(normal_rf,xlab = "Normalized RF distance", main = NULL)


normal_rf_df <- data_frame(gene=names(normal_rf), distance = normal_rf)

### number of tips in the gene trees
number_tips <- c()
for (i in seq(length(locus.trees))) {
  len <- length(locus.trees[[i]]$tip.label)
  number_tips <- c(number_tips,len)
}
table(number_tips)
hist(number_tips, density = T)
```

```{r}
TreeDist::TreeDistance(species_tree,locus.trees[1])

normal_treedist <- c()
for (i in seq(length(locus.trees))) {
  rf <- TreeDist::TreeDistance(species_tree,locus.trees[i])
  normal_treedist <- c(normal_treedist,rf)
}
names(normal_treedist) <- og_list$V1
normal_treedist_df <- data_frame(gene=names(normal_treedist), distance = normal_treedist)

```

```{r}
#### 99%
genes_99_treedist <- normal_treedist[which(normal_treedist > quantile(normal_treedist,0.99))]
names(genes_99_treedist) ### 53 genes

normal_treedist_df %>% 
  filter(gene %in% c(names(genes_99_treedist))) %>% 
  write_tsv("/Users/hsf378/PhD/Analysis/turtle-jeans/Results/species_vs_gene_treedist_99%_outiler.tsv")
#### 95%
genes_95_treedist <- normal_treedist[which(normal_treedist > quantile(normal_treedist,0.95))]
names(genes_95_treedist) ### 262 genes

normal_treedist_df %>% 
  filter(gene %in% c(names(genes_95_treedist))) %>% 
  write_tsv("/Users/hsf378/PhD/Analysis/turtle-jeans/Results/species_vs_gene_treedist_95%_outiler.tsv")

#### 90%
genes_90_treedist <- normal_treedist[which(normal_treedist > quantile(normal_treedist,0.9))]
names(genes_90_treedist) ### 524 genes

normal_treedist_df %>% 
  filter(gene %in% c(names(genes_90_treedist))) %>% 
  write_tsv("/Users/hsf378/PhD/Analysis/turtle-jeans/Results/species_vs_gene_treedist_90%_outiler.tsv")
```



### outlier genes

#### outlier genes with 99% confidence interval
```{r}
genes_99 <- normal_rf[which(normal_rf > quantile(normal_rf,0.99))]
names(genes_99) ### 52 genes

normal_rf_df %>% 
  filter(gene %in% c(names(genes_99))) %>% 
  write_tsv("/Users/hsf378/PhD/Analysis/turtle-jeans/Results/species_vs_gene_99%_outiler.tsv")

```

##### visualization
```{r}
outlier_colour <- "#727149FF"
no_outlier_colour <- "#A2BAC5FF"
outliers_cutoff <- quantile(normal_rf,0.99)
density_data <- density(normal_rf_df$distance)
density_data_df <- tibble(distance=density_data$x,
                          density=density_data$y)
density_data_df <- density_data_df %>% 
  mutate(outlier = case_when(distance > outliers_cutoff ~ "Outlier",
                             TRUE ~ "No outlier"))

ggplot(density_data_df, aes(x=distance, y=density, fill=outlier)) +
  geom_line(color = no_outlier_colour) +
  geom_ribbon(aes(ymin = 0, ymax = density), alpha = 0.8) +
  scale_fill_manual("Gene category",
                    values = c("No outlier" = no_outlier_colour, "Outlier" = outlier_colour)) + 
  geom_vline(xintercept = outliers_cutoff, linetype = "dashed", 
             color = "grey40", size = 0.4) +
  annotate("text", x = outliers_cutoff + 0.01, y = 0.3,  # Position text slightly below the peak
           label = "99% Cutoff", color = "grey40", hjust = -0.1, size=4) +
  xlab("Normalized RF distance") +
  ylab("Density") +
  theme_minimal()
ggsave("/Users/hsf378/PhD/Analysis/turtle-jeans/Results/species_vs_gene_99_density.pdf", width=7, height=5)
```


#### outlier genes with 95% confidence interval
```{r}
genes_95 <- normal_rf[which(normal_rf > quantile(normal_rf,0.95))]
names(genes_95) ### 249 genes

normal_rf_df %>% 
  filter(gene %in% c(names(genes_95))) %>% 
  write_tsv("/Users/hsf378/PhD/Analysis/turtle-jeans/Results/species_vs_gene_95%_outiler.tsv")
```


```{r}
outlier_colour <- "#727149FF"
no_outlier_colour <- "#A2BAC5FF"
outliers_cutoff <- quantile(normal_rf,0.95)
density_data <- density(normal_rf_df$distance)
density_data_df <- tibble(distance=density_data$x,
                          density=density_data$y)
density_data_df <- density_data_df %>% 
  mutate(outlier = case_when(distance > outliers_cutoff ~ "Outlier",
                             TRUE ~ "No outlier"))

ggplot(density_data_df, aes(x=distance, y=density, fill=outlier)) +
  geom_line(color = no_outlier_colour) +
  geom_ribbon(aes(ymin = 0, ymax = density), alpha = 0.8) +
  scale_fill_manual("Gene category",
                    values = c("No outlier" = no_outlier_colour, "Outlier" = outlier_colour)) + 
  geom_vline(xintercept = outliers_cutoff, linetype = "dashed", 
             color = "grey40", size = 0.4) +
  annotate("text", x = outliers_cutoff + 0.01, y = 0.3,  # Position text slightly below the peak
           label = "95% Cutoff", color = "grey40", hjust = -0.1, size=4) +
  xlab("Normalized RF distance") +
  ylab("Density") +
  theme_minimal()
ggsave("/Users/hsf378/PhD/Analysis/turtle-jeans/Results/species_vs_gene_95_density.pdf", width=7, height=5)
```


#### outlier genes with 90% confidence interval
```{r}
genes_90 <- normal_rf[which(normal_rf > quantile(normal_rf,0.90))]
names(genes_90) ### 502 genes

normal_rf_df %>% 
  filter(gene %in% c(names(genes_90))) %>% 
  write_tsv("/Users/hsf378/PhD/Analysis/turtle-jeans/Results/species_vs_gene_90%_outiler.tsv")
```

```{r}
outlier_colour <- "#727149FF"
no_outlier_colour <- "#A2BAC5FF"
outliers_cutoff <- quantile(normal_rf,0.90)
density_data <- density(normal_rf_df$distance)
density_data_df <- tibble(distance=density_data$x,
                          density=density_data$y)
density_data_df <- density_data_df %>% 
  mutate(outlier = case_when(distance > outliers_cutoff ~ "Outlier",
                             TRUE ~ "No outlier"))

ggplot(density_data_df, aes(x=distance, y=density, fill=outlier)) +
  geom_line(color = no_outlier_colour) +
  geom_ribbon(aes(ymin = 0, ymax = density), alpha = 0.8) +
  scale_fill_manual("Gene category",
                    values = c("No outlier" = no_outlier_colour, "Outlier" = outlier_colour)) + 
  geom_vline(xintercept = outliers_cutoff, linetype = "dashed", 
             color = "grey40", size = 0.4) +
  annotate("text", x = outliers_cutoff + 0.01, y = 0.3,  # Position text slightly below the peak
           label = "90% Cutoff", color = "grey40", hjust = -0.1, size=4) +
  xlab("Normalized RF distance") +
  ylab("Density") +
  theme_minimal()
ggsave("/Users/hsf378/PhD/Analysis/turtle-jeans/Results/species_vs_gene_90_density.pdf", width=7, height=5)
```


```{r}
# Find the indices where RF equal to 1
high_rf_indices <- which(normal_rf == 1)

# Extract the RF distances that are greater than 0.9
high_rf_distances <- normal_rf[high_rf_indices]

og_candidate <- og_list[high_rf_indices,]
og_candidate_1 <- og_candidate[which(Ntip(locus.trees[high_rf_indices])>10)]
```


### perfect gene trees
```{r}
confident_gene_tree <- names(which(normal_rf == 0))

confident_genes <- confident_gene_tree[which(Ntip(locus.trees[which(normal_rf == 0)])>10)]
```


### Densitree

```{r}
### extract outlier gene trees 
locus.trees[which(names %in% og_candidate_1)]
# plot(locus.trees[which(names %in% og_candidate_1)])
# Combine the species tree and gene trees into one list
merged_trees <- c(species_tree, locus.trees[which(names %in% og_candidate_1)])
p_densi_disagree <- densiTree(merged_trees, col = c("black", rep('steelblue', 7)), alpha = 0.7)
p_hist <- hist(normal_rf,xlab = "Normalized RF distance", main = NULL)

# plot(merge_species_confidence)
merge_species_confidence <- c(locus.trees[which(names %in% confident_genes)], species_tree)
p_densi_agree <- phangorn::densiTree(merge_species_confidence, col = c( rep('steelblue', 216),"black"), alpha = 0.7)


ggarrange(p_densi_agree,p_densi_disagree, p_hist,
          labels = c("A", "B","C"),
          nrow = 3)

pdf("densiTree_plot.pdf", width = 10, height = 7)
png("densiTree_plot.png", width = 1400, height = 700, res = 150)
par(mfrow = c(1, 2))
#phangorn::densiTree(merge_species_confidence, col = c( rep('steelblue', 216),"black"), alpha = 0.7)
hist(normal_rf,xlab = "Normalized RF distance", main = NULL)
densiTree(merged_trees, col = c("black", rep('steelblue', 7)), alpha = 0.7, use.edge.length = F, width = 2,scaleX = 1)
dev.off
```


### GO enrichment analysis

```{r}
### turtles annotation db
go_all <- read.table("/Users/hsf378/Downloads/candidate_at32523.txt", header = F, col.names = c("GID","ontology","GO","Support_number"), sep = "\t")

### subset DB associated with biological process
go_bp <- go_all %>% 
  filter(ontology == "biological_process",
         str_detect(GO,"GO")) %>% 
  select(GO,GID) %>% 
  mutate(goterm = Term(GO))
go_bp_name <- go_bp %>% select(GO,GID)
go_bp_term <- go_bp %>% select(GO,goterm)

### subset DB associated with cellular component
go_cc <- go_all %>% 
  filter(ontology == "cellular_component",
         str_detect(GO,"GO")) %>% 
  select(GO,GID) %>% 
  mutate(goterm = Term(GO))
go_cc_name <- go_cc %>% select(GO,GID)
go_cc_term <- go_cc %>% select(GO,goterm)


### subset DB associated with molecular function
go_mf <- go_all %>% 
  filter(ontology == "molecular_function",
         str_detect(GO,"GO")) %>% 
  select(GO,GID) %>% 
  mutate(goterm = Term(GO))
go_mf_name <- go_mf %>% select(GO,GID)
go_mf_term <- go_mf %>% select(GO,goterm)


### DB with all ontologies

go_all_name <- go_all %>% filter(str_detect(GO,"GO")) %>% select(GO,GID)
go_all_term <- go_all %>% filter(str_detect(GO,"GO")) %>% select(GO,GID) %>% mutate(goterm = Term(GO)) %>% select(GO,goterm)
# all_enrich <- enricher(gene = og_candidate_1,
#                         # pvalueCutoff = 100,
#                         pAdjustMethod = "fdr",
#                         # qvalueCutoff = 100,
#                         TERM2GENE = go_all_name,
#                         TERM2NAME = go_all_term)

bp_enrich <- enricher(gene = og_candidate_1,
                        # pvalueCutoff = 100,
                        pAdjustMethod = "fdr",
                        # qvalueCutoff = 100,
                        TERM2GENE = go_bp_name,
                        TERM2NAME = go_bp_term)
cc_enrich <- enricher(gene = og_candidate_1,
                        # pvalueCutoff = 100,
                        pAdjustMethod = "fdr",
                        # qvalueCutoff = 100,
                        TERM2GENE = go_cc_name,
                        TERM2NAME = go_cc_term)
mf_enrich <- enricher(gene = og_candidate_1,
                        # pvalueCutoff = 100,
                        pAdjustMethod = "fdr",
                        # qvalueCutoff = 100,
                        TERM2GENE = go_mf_name,
                        TERM2NAME = go_mf_term)

### visualization

# Add an 'ontology' column to each data frame to indicate the GO category
bp_df <- as.data.frame(bp_enrich) %>% mutate(ontology = "BP")
cc_df <- as.data.frame(cc_enrich) %>% mutate(ontology = "CC")
mf_df <- as.data.frame(mf_enrich) %>% mutate(ontology = "MF")

# Combine the data frames into one
combined_result <- bind_rows(bp_df, cc_df, mf_df)

# Select only necessary columns for plotting (for example, "Description", "Count", and "ontology")
# Make sure to adjust column names if different in your enricher output
plot_data <- combined_result %>%
  select(Description, Count,p.adjust, ontology) %>% 
  remove_missing(.)

# Create the bar plot with facet_wrap
ggplot(plot_data, aes(x = Description, y = Count, fill = p.adjust)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~ ontology, scales = "free_y",nrow = 3) +
  theme_bw() +
  xlab(NULL) +
  ylab("Count") +
  scale_fill_gradient2(low = "grey", mid = "yellow", high = "red") +
  theme(strip.text = element_text(size = 12),  # Adjust facet label size
        axis.text = element_text(size = 10),   # Adjust axis text size
        axis.title = element_text(size = 14))  # Adjust axis title size

ggsave("./Goenrich_speciesvsgenes.png", height = 7, width = 9)

# GSEA(gene = normal_rf[og_candidate_1][order(normal_rf[og_candidate_1], decreasing = T)],
#                         # pvalueCutoff = 100,
#                         pAdjustMethod = "fdr",
#                         # qvalueCutoff = 100,
#                         TERM2GENE = go_all_name,
#                         TERM2NAME = go_all_term)
```

## gene trees vs gene trees

### gene trees vs gene trees
```{r}
# load gene trees
confident_trees <- locus.trees[Ntip(locus.trees)>10]
# load gene names
confident_names <- names[Ntip(locus.trees)>10]

## metadata for turtles
meta_turtles <- read_tsv("Data/metadata_habitat.tsv")

results <- phylter(confident_trees, gene.names = confident_names, k=6)

# OUTLIERS
outliers_og_subset <- results$Final$Outliers[,1]

names(outliers_og_subset) <- results$Final$Outliers[,2]

table(results$Final$Outliers[,2])

unique(outliers_og_subset)


Outliers_all <- results$Final$Outliers %>% 
  as_data_frame(.) %>% 
  left_join(.,meta_turtles, by = c("V2" = "Species")) %>% 
  remove_missing(.)

# Outliers_Freshwater <- Outliers_all %>% filter(Habitat == "Freshwater")
Outliers_Rivers <- Outliers_all %>% filter(Habitat == "Rivers")
Outliers_Swamps <- Outliers_all %>% filter(Habitat == "Swamps")
Outliers_Lakes <- Outliers_all %>% filter(Habitat == "Lakes")
Outliers_Terrestrial<- Outliers_all %>% filter(Habitat == "Terrestrial")
Outliers_Marine <- Outliers_all %>% filter(Habitat == "Marine")

```


### gene enrichment analysis

```{r}
enrich_plot <- function(df,plot_name,h,w){
  bp <- enricher(gene = unique(df$V1),
                        pAdjustMethod = "fdr",
                        TERM2GENE = go_bp_name,
                        TERM2NAME = go_bp_term)
  cc <- enricher(gene = unique(df$V1),
                        pAdjustMethod = "fdr",
                        TERM2GENE = go_cc_name,
                        TERM2NAME = go_cc_term)
  mf <- enricher(gene = unique(df$V1),
                        pAdjustMethod = "fdr",
                        TERM2GENE = go_mf_name,
                        TERM2NAME = go_mf_term)
  # Add an 'ontology' column to each data frame to indicate the GO category
  bp_df1 <- as.data.frame(bp) %>% mutate(ontology = "BP")
  cc_df1 <- as.data.frame(cc) %>% mutate(ontology = "CC")
  mf_df1 <- as.data.frame(mf) %>% mutate(ontology = "MF")

  # Combine the data frames into one
  combined_result1 <- bind_rows(bp_df1, cc_df1, mf_df1)

  # Select only necessary columns for plotting (for example, "Description", "p.adjust", "Count", and "ontology")
  plot_data1 <- combined_result1 %>%
    select(Description, Count, p.adjust, ontology) %>% 
    remove_missing(.)

  # Create the bar plot with facet_wrap
  ggplot(plot_data1, aes(x = Description, y = Count, fill = p.adjust)) +
    geom_bar(stat = "identity") +
    coord_flip() +
    facet_wrap(~ ontology, scales = "free_y",nrow = 3) +
    theme_bw() +
    xlab(NULL) +
    ylab("Count") +
    scale_fill_gradient2(low = "grey", mid = "yellow", high = "red") +
    theme(strip.text = element_text(size = 12),  # Adjust facet label size
        axis.text = element_text(size = 10),   # Adjust axis text size
        axis.title = element_text(size = 14))  # Adjust axis title size

  ggsave(plot_name, height = h, width = w)

  
}
# enrich_plot(Outliers_Rivers,"Outliers_Rivers.png",10,9)
enrich_plot(Outliers_Swamps,"Outliers_Swamps.png",10,9)
enrich_plot(Outliers_Lakes,"Outliers_Lakes.png",10,9)
enrich_plot(Outliers_Marine,"Outliers_Marine.png",10,9)
# enrich_plot(Outliers_Terrestrial,"Outliers_Terrestrial.png",10,9)

```


### DensiTree
```{r}
merge_species_gene_trees <- c( locus.trees[which(names %in% unique(outliers_og_subset))], species_tree)

col_all_tree <- c(rep('steelblue', 47),"black")
pdf("densiGeneTree_plot.pdf", width = 10, height = 7)
phangorn::densiTree(merge_species_gene_trees, col = col_all_tree, alpha = 0.7, width = 1,use.edge.length = FALSE, adj = 0.5, scale.bar = F)
dev.off
```
### Overlap outlier genes in different habitat 
```{r}
list_genes <- list(Lake = unique(Outliers_Lakes$V1),
                   Swamp = unique(Outliers_Swamps$V1),
                   River = unique(Outliers_Rivers$V1),
                   Marine = unique(Outliers_Marine$V1),
                   Terrestrial = unique(Outliers_Terrestrial$V1))

m = make_comb_mat(list_genes)
png("outlier_gene_habitat.png", width = 1000, height = 700, res = 150)
UpSet(m, top_annotation = upset_top_annotation(m, add_numbers = TRUE),
    right_annotation = upset_right_annotation(m, add_numbers = TRUE))
dev.off()

```


